
---
title: "Traffic Accident Data Analysis"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
```

## Load Data

```{r load-data}
# Load all datasets
accident <- read_csv("accident.csv")
vehicle <- read_csv("vehicle.csv")
person <- read_csv("person.csv")
race <- read_csv("race.csv")
weather <- read_csv("weather.csv")
damage <- read_csv("damage.csv")
safetyeq <- read_csv("safetyeq.csv")
vsoe <- read_csv("vsoe.csv")
nmcrash <- read_csv("nmcrash.csv")

# Merge datasets
merged_data <- person %>%
  left_join(vehicle, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(accident, by = "ST_CASE") %>%
  left_join(weather, by = "ST_CASE") %>%
  left_join(race, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  left_join(safetyeq, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  left_join(damage, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(vsoe, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(nmcrash, by = c("ST_CASE", "VEH_NO", "PER_NO"))
```

## 👤 Person-Based Analysis

### 1. What age group has the highest fatality rate?

```{r age-fatality}
ggplot(merged_data, aes(x = as.numeric(AGE), fill = INJ_SEVNAME)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.6) +
  labs(title = "Age vs. Injury Severity", x = "Age", y = "Number of People")
```
*This plot shows how injury severity varies by age. We can visually identify if fatal injuries are more frequent in a specific age group.*

### 2. Does seat position affect injury severity?

```{r seat-position}
ggplot(merged_data, aes(x = SEAT_POSNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Injury Severity by Seat Position", x = "Seat Position", y = "Proportion")
```
*This bar chart compares seat position with injury severity. Rear seats, front seats, and driver seats can be compared for safety insights.*

### 3. Are males or females more frequently involved?

```{r gender-involvement}
ggplot(merged_data, aes(x = SEXNAME)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Number of People Involved by Gender", x = "Gender", y = "Count")
```
*This chart simply shows the count of males and females involved in crashes.*

## 🚗 Vehicle-Based Analysis

### 4. Which body types are involved in the most crashes?

```{r body-type}
ggplot(merged_data, aes(x = BODY_TYPNAME)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "Crashes by Vehicle Body Type", x = "Body Type", y = "Count")
```
*Shows which vehicle body types are most commonly involved in accidents.*

### 5. Is there a link between vehicle weight and injuries?

```{r weight-injury}
ggplot(merged_data, aes(x = GVWR_FROMNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Vehicle Weight Class vs. Injury Severity", x = "Weight Class", y = "Proportion")
```
*This plot helps examine whether heavier or lighter vehicles result in more severe injuries.*

### 6. Are rollovers more likely to result in fatalities?

```{r rollover-fatality}
ggplot(merged_data, aes(x = ROLLOVERNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  labs(title = "Rollover Involvement and Injury Severity", x = "Rollover", y = "Proportion")
```
*Highlights whether rollover accidents lead to more fatal outcomes.*

## 🌦️ Weather-Based Analysis

### 7. In which weather condition do most accidents happen?

```{r weather-accidents}
ggplot(merged_data, aes(x = WEATHERNAME)) +
  geom_bar(fill = "tomato") +
  coord_flip() +
  labs(title = "Accidents by Weather Condition", x = "Weather", y = "Count")
```
*Shows which weather conditions are associated with the most crashes.*

### 8. Are snowy or rainy days more dangerous?

```{r weather-severity}
ggplot(merged_data, aes(x = WEATHERNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Weather Condition vs. Injury Severity", x = "Weather", y = "Proportion")
```
*Compares different weather types and their impact on injury severity.*

## 🧯 Safety Equipment Analysis

### 9. Did seatbelt or helmet usage reduce fatalities?

```{r restraint-fatality}
ggplot(merged_data, aes(x = REST_USENAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Restraint Use vs. Injury Severity", x = "Restraint Use", y = "Proportion")
```
*Analyzes how using seatbelts correlates with reduced fatalities.*

### 10. Was airbag deployment associated with survival?

```{r airbag-survival}
ggplot(merged_data, aes(x = AIR_BAGNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Airbag Deployment and Injury Severity", x = "Airbag Deployment", y = "Proportion")
```
*Visualizes the relationship between airbag deployment and injury outcome.*

