---
title: "Traffic Accident Data Analysis with Preprocessing"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)

```

## 1. Load Data

```{r load-data}
# Read datasets
accident <- read_csv("accident.csv")
vehicle <- read_csv("vehicle.csv")
person <- read_csv("person.csv")
race <- read_csv("race.csv")
weather <- read_csv("weather.csv")
damage <- read_csv("damage.csv")
safetyeq <- read_csv("safetyeq.csv")
vsoe <- read_csv("vsoe.csv")
nmcrash <- read_csv("nmcrash.csv")
```

## 2. Preprocessing and Cleaning

```{r preprocessing}
# Remove duplicates if any

person <- person %>%
  distinct() %>%
  mutate(AGE = as.numeric(AGE)) %>%
  drop_na()

vehicle <- vehicle %>%
  distinct() %>%
  drop_na()

accident <- accident %>%
  distinct() %>%
  drop_na()

weather <- weather %>%
  distinct() %>%
  drop_na()

race <- race %>%
  distinct() %>%
  drop_na()

damage <- damage %>%
  distinct() %>%
  drop_na()

safetyeq <- safetyeq %>%
  distinct() %>%
  drop_na()

vsoe <- vsoe %>%
  distinct() %>%
  drop_na()

nmcrash <- nmcrash %>%
  distinct() %>%
  drop_na()
# Handle missing values (basic strategy: filter or replace with 'Unknown')
person <- person %>%
  mutate(
    AGE = as.numeric(AGE),
    INJ_SEVNAME = replace_na(INJ_SEVNAME, "Unknown"),
    SEXNAME = replace_na(SEXNAME, "Unknown"),
    SEAT_POSNAME = replace_na(SEAT_POSNAME, "Unknown")
  )

vehicle <- vehicle %>%
  mutate(
    BODY_TYPNAME = replace_na(BODY_TYPNAME, "Unknown"),
    GVWR_FROMNAME = replace_na(GVWR_FROMNAME, "Unknown"),
    ROLLOVERNAME = replace_na(ROLLOVERNAME, "Unknown")
  )

weather <- weather %>%
  mutate(WEATHERNAME = replace_na(WEATHERNAME, "Unknown"))

# Convert time columns to numeric
person <- person %>%
  mutate(HOUR = as.numeric(HOUR), AGE = as.numeric(AGE))
```

## 3. Merge Datasets

```{r merging}
# Merge data into one master dataset
merged_data <- person %>%
  left_join(vehicle, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(accident, by = "ST_CASE") %>%
  left_join(weather, by = "ST_CASE") %>%
  left_join(race, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  left_join(safetyeq, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  left_join(damage, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(vsoe, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(nmcrash, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  filter(!is.na(ST_CASE))

colnames(merged_data)
```

## 4. Exploratory Data Analysis

### 👤 Person-Based Questions

#### Age group with the highest fatality rate

```{r}

filtered_data <- merged_data %>%
  filter(AGE < 100)

ggplot(filtered_data, aes(x = AGE, fill = INJ_SEVNAME)) +
  geom_histogram(binwidth = 5, alpha = 0.7) +
  labs(title = "Age vs Injury Severity", x = "Age", y = "Count")
```



```{r}
filtered_data %>%
  count(AGE, sort = TRUE) %>%
  slice_max(n, n = 4)
```


#### Seat position vs. injury severity

```{r}

ggplot(merged_data, aes(x = SEAT_POSNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Seat Position vs Injury Severity", x = "Seat Position", y = "Proportion")

```

#### Gender involvement

```{r}

ggplot(merged_data, aes(x = SEXNAME)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Gender Involvement in Crashes", x = "Gender", y = "Count")
```


### 🚗 Vehicle-Based Questions

#### Vehicle body types involved

```{r}
ggplot(merged_data, aes(x = BODY_TYP.x)) +
  geom_bar(fill = "steelblue") +
  coord_flip() +
  labs(title = "Crashes by Vehicle Body Type", x = "Vehicle Body Type", y = "Count")
unique(merged_data$VE_FORMS.x)
length(unique(merged_data$VPICBODYCLASSNAME.x))
unique(merged_data$GVWR_FROMNAME.x)
```

```{r}
merged_data %>%
  count(BODY_TYPNAME.x, sort = TRUE) %>%
  slice_max(n, n = 4)
```

#### Vehicle weight class vs. injury

```{r weight-injury}

new_df <- merged_data %>%
  select(GVWR_FROMNAME.x, INJ_SEVNAME) %>%
  filter(!is.na(GVWR_FROMNAME.x))


ggplot(new_df, aes(x = GVWR_FROMNAME.x, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Vehicle Weight Class vs. Injury Severity", x = "Weight Class", y = "Proportion")

```
#### Rollover vs. fatalities

```{r}

ggplot(merged_data, aes(x = ROLLOVERNAME.x, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  labs(
    title = "Rollover Involvement and Injury Severity",
    x = "Rollover Status",
    y = "Proportion"
  ) +
  coord_flip()

```

```{r}
rollover_data %>%
  count(ROLLOVERNAME.x, sort = TRUE) %>%
  slice_max(n, n = 4)

```
### Substance Involvement Combo Analysis

```{r}
substance_data <- merged_data %>%
  filter(!is.na(DRINKINGNAME), !is.na(DRUGSNAME), !is.na(INJ_SEVNAME)) %>%
  mutate(
    substance_combo = case_when(
      DRINKINGNAME == "Yes" & DRUGSNAME == "Yes" ~ "Alcohol & Drugs",
      DRINKINGNAME == "Yes" & DRUGSNAME != "Yes" ~ "Alcohol Only",
      DRINKINGNAME != "Yes" & DRUGSNAME == "Yes" ~ "Drugs Only",
      TRUE ~ "None"
    )
  )
ggplot(substance_data, aes(x = substance_combo, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  labs(
    title = "Injury Severity by Substance Involvement",
    x = "Substance Use Category",
    y = "Proportion"
  ) +
  coord_flip()


```

### Location Based

### State-Based Analysis
### Which states have the most crashes?
```{r}
state_crashes <- merged_data %>%
  filter(!is.na(STATENAME)) %>%
  count(STATENAME, sort = TRUE)

ggplot(state_crashes, aes(x = reorder(STATENAME, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Total Crashes by State", x = "State", y = "Crash Count")


```
### Which states have the highest fatality rates?

```{r}
fatal_by_state <- merged_data %>%
  filter(!is.na(STATENAME), !is.na(INJ_SEVNAME)) %>%
  group_by(STATENAME) %>%
  summarise(
    total = n(),
    fatal = sum(INJ_SEVNAME == "Fatal Injury (K)"),
    fatal_rate = fatal / total * 100
  ) %>%
  arrange(desc(fatal_rate))

ggplot(fatal_by_state, aes(x = reorder(STATENAME, fatal_rate), y = fatal_rate)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(title = "Fatality Rate by State", x = "State", y = "Fatality Rate (%)")
```




```{r}
fatal_by_state %>%
  count(STATENAME, sort = TRUE) %>%
  slice_max(fatal_rate, n = 4)

```

### Are rural or urban crashes more severe?
```{r}
ggplot(merged_data, aes(x = RUR_URBNAME.x, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  labs(
    title = "Injury Severity in Rural vs Urban Crashes",
    x = "Location Type",
    y = "Proportion"
  ) +
  coord_flip()
```

### 🌦 Weather-Based Questions

#### Weather with most accidents

```{r}

ggplot(merged_data, aes(x = WEATHERNAME.x)) +
  geom_bar(fill = "orange") +
  coord_flip() +
  labs(title = "Accidents by Weather Condition", x = "Weather", y = "Count")

```

#### Snow/Rain and injury severity

```{r}

ggplot(weather_data, aes(x = WEATHERNAME.x, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Weather vs Injury Severity", x = "Weather", y = "Proportion")

### Time based
```

```{r}

library(dplyr)
library(ggplot2)

merged_data <- merged_data %>%
  filter(!is.na(HOUR), HOUR < 24) %>%
  mutate(
    TIME_OF_DAY = case_when(
      HOUR >= 0  & HOUR <= 5  ~ "Night",
      HOUR >= 6  & HOUR <= 11 ~ "Morning",
      HOUR >= 12 & HOUR <= 16 ~ "Afternoon",
      HOUR >= 17 & HOUR <= 20 ~ "Evening",
      HOUR >= 21 & HOUR <= 23 ~ "Late Night"
    )
  )

heatmap_data <- merged_data %>%
  filter(!is.na(INJ_SEVNAME)) %>%
  group_by(TIME_OF_DAY, INJ_SEVNAME) %>%
  summarise(count = n(), .groups = "drop")

ggplot(heatmap_data, aes(x = TIME_OF_DAY, y = INJ_SEVNAME, fill = count)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "lightyellow", high = "darkred") +
  labs(
    title = "Heatmap of Injury Severity by Time of Day",
    x = "Time of Day",
    y = "Injury Severity",
    fill = "Crash Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    panel.grid = element_blank()
  )


```


### 🧯 Safety Equipment Questions

#### Seatbelt or helmet use vs. fatalities

```{r}

ggplot(safety_data, aes(x = REST_USENAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Seatbelt Use vs. Injury Severity", x = "Seatbelt Usage", y = "Proportion")

```

#### Airbag deployment and survival

```{r}
ggplot(merged_data, aes(x = AIR_BAGNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Airbag Deployment vs. Injury Severity", x = "Airbag Status", y = "Proportion")

```
## ✅ Final Summary

This report provides a complete picture of how personal, vehicle, and environmental factors contribute to crash outcomes. The preprocessing ensures cleaner data and more accurate visual insights.