---
title: "PROJECT"
author: "Kirubhaharan"
date: "2025-04-16"
output: pdf_document
---

```{r}

library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(lubridate)

```


# 1. Load Data

```{r}

accident <- read_csv("accident.csv")
vehicle <- read_csv("vehicle.csv")
person <- read_csv("person.csv")
race <- read_csv("race.csv")
weather <- read_csv("weather.csv")
damage <- read_csv("damage.csv")
safetyeq <- read_csv("safetyeq.csv")
vsoe <- read_csv("vsoe.csv")
nmcrash <- read_csv("nmcrash.csv")

```

# 2. Preprocessing and Cleaning

```{r}

# Limit to 20000 rows to prevent memory crash
person <- person %>%
  distinct() %>%
  mutate(AGE = as.numeric(AGE)) %>%
  drop_na(AGE, INJ_SEVNAME)


vehicle <- vehicle %>%
  distinct() %>%
  drop_na()

accident <- accident %>%
  distinct() %>%
  drop_na()

weather <- weather %>%
  distinct() %>%
  drop_na()

race <- race %>%
  distinct() %>%
  drop_na()

damage <- damage %>%
  distinct() %>%
  drop_na()

safetyeq <- safetyeq %>%
  distinct() %>%
  drop_na()

vsoe <- vsoe %>%
  distinct() %>%
  drop_na()

nmcrash <- nmcrash %>%
  distinct() %>%
  drop_na()
# Handle missing values (basic strategy: filter or replace with 'Unknown')
# Limit to 20000 rows to prevent memory crash
person <- person %>%
  mutate(
    AGE = as.numeric(AGE),
    INJ_SEVNAME = replace_na(INJ_SEVNAME, "Unknown"),
    SEXNAME = replace_na(SEXNAME, "Unknown"),
    SEAT_POSNAME = replace_na(SEAT_POSNAME, "Unknown")
  )

vehicle <- vehicle %>%
  mutate(
    BODY_TYPNAME = replace_na(BODY_TYPNAME, "Unknown"),
    GVWR_FROMNAME = replace_na(GVWR_FROMNAME, "Unknown"),
    ROLLOVERNAME = replace_na(ROLLOVERNAME, "Unknown")
  )

weather <- weather %>%
  mutate(WEATHERNAME = replace_na(WEATHERNAME, "Unknown"))

# Convert time columns to numeric
# Limit to 20000 rows to prevent memory crash
person <- person %>%
  mutate(HOUR = as.numeric(HOUR), AGE = as.numeric(AGE))

```

# 3. Merge Datasets

```{r}

merged_data <- suppressWarnings(person %>%
  left_join(vehicle, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(accident, by = "ST_CASE") %>%
  left_join(weather, by = "ST_CASE") %>%
  left_join(race, by = c("ST_CASE", "VEH_NO", "PER_NO"))%>%
  left_join(safetyeq, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  left_join(damage, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(vsoe, by = c("ST_CASE", "VEH_NO")) %>%
  left_join(nmcrash, by = c("ST_CASE", "VEH_NO", "PER_NO")) %>%
  filter(!is.na(ST_CASE)))

```

# 4. Exploratory Data Analysis

#  Person-Based Questions

# Age group with the highest fatality rate

```{r}

library(ggplot2)

filtered_data <- merged_data %>%
  filter(AGE < 100)

ggplot(filtered_data, aes(x = AGE, fill = INJ_SEVNAME)) +
  geom_histogram(binwidth = 5, alpha = 0.7) +
  labs(title = "Age vs Injury Severity", x = "Age", y = "Count")

```

```{r}
filtered_data %>%
  count(AGE, sort = TRUE) %>%
  slice_max(n, n = 5)

```

```{r}

library(ggplot2)
fatal_age_data <- merged_data %>%
  filter(AGE < 100, INJ_SEVNAME == "Fatal Injury (K)")

ggplot(fatal_age_data, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "firebrick", alpha = 0.8) +
  labs(title = "Age Distribution of Fatal Injuries", x = "Age", y = "Fatal Crash Count")

```


```{r}
fatal_age_data %>%
  count(AGE, sort = TRUE) %>%
  slice_max(n, n = 5)

```
The visualization and table highlights that the age 19 has the highest fatal injury rate and the highest number of fatal crash victims are in the late teen to mid-20s age range. This suggests that young adult drivers (especially those aged 18–24) are disproportionately represented in fatal crash statistics.

# Seat position vs injury severity

```{r}
library(ggplot2)
merged_data <- merged_data %>%
  mutate(SEAT_POSNAME = case_when(
    SEAT_POSNAME == "Other Passenger in enclosed passenger or cargo area" ~ "Other - Enclosed",
    SEAT_POSNAME == "Other Passenger in passenger or cargo area, unknown whether or not enclosed" ~ "Other - Unknown",
    SEAT_POSNAME == "Other Passenger in unenclosed passenger or cargo area" ~ "Other - Unenclosed",
    TRUE ~ SEAT_POSNAME
  ))

ggplot(merged_data, aes(x = SEAT_POSNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Seat Position vs Injury Severity", x = "Seat Position", y = "Proportion")

```
# Major seat position with injuries

```{r}

library(ggplot2)

major_seats <- merged_data %>%
  filter(
    SEAT_POSNAME %in% c(
      "Front Seat, Left Side",      # Driver
      "Front Seat, Right Side",     # Front passenger
      "Second Seat, Left Side",
      "Second Seat, Right Side",
      "Second Seat, Middle"),
    !is.na(INJ_SEVNAME)
  )
ggplot(major_seats, aes(x = SEAT_POSNAME, fill = INJ_SEVNAME)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(
    title = "Injury Severity by Major Seat Positions",
    x = "Seat Position",
    y = "Proportion")

```

From this visualization Front seat,Left side which is the driver seat is having more injury rate.